---
title: "Balance Test"
author: "Satoshi Koiso"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: 
  html_document:
    theme: "readable"
    toc: true
    toc_float: true
  word_document: default
  pdf_document:
    df_print: paged
    latex_engine: xelatex
always_allow_html: true
---

```{r setup, include=FALSE}
if (TRUE) {
  list.of.packages <- c("tidyverse", "haven", "here",
                      "kableExtra","rstudioapi","DescTools","qwraps2","data.table",
                      "stargazer","readxl","dplyr","rootSolve","ICC", "randomizr",
                      "multiwayvcov", "lmtest", "knitr","devtools", "Hmisc",
                      "compareGroups", "naniar")
  # sapply(list.of.packages, function(x) pacman::p_load(get(x)))

  new.packages <- list.of.packages[
    !( list.of.packages %in% installed.packages()[,"Package"] )
    ]
  if(length(new.packages)) install.packages(new.packages,
                                            repos = "https://cloud.r-project.org")

  lapply(list.of.packages, library, character.only = TRUE)
}

devtools::install_github('vikjam/pwrcalc')
## If you have trouble installing pwrcalc, you can download it manually by using the .tar.gz file provided. 
## To install it manually, go to "Packages", select "Install" and then "Package Archive" in the install menu. 

setwd(dirname(getActiveDocumentContext()$path)) 
knitr::opts_knit$set(root.dir = here())
knitr::opts_chunk$set(echo = T, eval = T, message = F, warning = F, fig.pos = "H", fig.height = 7, fig.width = 7)

print_code <- TRUE
```


# Balance Test
 
The aim of randomization is to produce a balance in characteristics in the treatment/control groups.


# Example 

The `balsakhi` dataset is provided with the package `pwrcalc`. We'll use this dataset to estimate the control mean:

```{r}
library(pwrcalc)
data(balsakhi)

# Let us view the `balsakhi `data that we have just loaded:
View(balsakhi)
```


# Data Exploration

Let's explore the balance in the `balsakhi` study by presenting the means and standard deviations of the baseline test scores (total, math, and verbal) and age, gender, division, research group, as well as whether they cheated, they were helped, and they took it later. This can be done very easily with the `compareGroups()` function. 


```{r}
#Generate summaries of the variables by treatment group and save results as baselines
baselines <- compareGroups(bal ~ age + gender + div + researchgroup + sessiond + pre_tot + pre_math + pre_verb + precheated+ prehelped + prelater + pre_totnorm + pre_mathnorm + pre_verbnorm, data = balsakhi)
#Use the createTable function to display the results saved in baselines
baseline.table <- createTable(baselines, show.ratio = FALSE, show.p.overall=TRUE)
#Display the created summary table
baseline.table

```


```{r}
# to show ratio in table,
createTable(baselines, show.ratio=TRUE)

# Odds Ratio (OR) is a measure of association between exposure and an outcome
```

It looks like some variables have missing data.
Let's do data cleaning.


```{r}
missing <- miss_var_summary(balsakhi)
print(missing, n=76)
# should not use age and takehome
# found empty cells ("") are not treated as NA

# Convert "" to NA
balsakhi[balsakhi == ""] <- NA

# chack NA
missing <- miss_var_summary(balsakhi)
print(missing, n=76)

# select columns
base_var <- c("bal","gender","div","researchgroup","sessiond","pre_tot","pre_math", "pre_verb","precheated","prehelped","prelater","pre_totnorm","pre_mathnorm","pre_verbnorm")
balsakhi_base <- balsakhi[,base_var]

# drop na
balsakhi_base_cleaned <- drop_na(balsakhi_base)
missing <- miss_var_summary(balsakhi_base_cleaned)
print(missing, n=76)
```

# Balance Table

Finally, create the balance table

```{r}
#Generate summaries of the variables by treatment group and save results as baselines
baselines <- compareGroups(bal ~ gender + div + researchgroup + sessiond + pre_tot + pre_math + pre_verb + precheated+ prehelped + prelater + pre_totnorm + pre_mathnorm + pre_verbnorm, data = balsakhi_base_cleaned)
#Use the createTable function to display the results saved in baselines
baseline.table <- createTable(baselines, show.ratio = FALSE, show.p.overall=TRUE)
#Display the created summary table
baseline.table

```


```{r}
# to show ratio in table,
createTable(baselines, show.ratio=TRUE)

# Odds Ratio (OR) is a measure of association between exposure and an outcome
```

Some variables are not balanced at baseline.