---
title: "Endpoint"
author: "Satoshi Koiso"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: 
  html_document:
    theme: "readable"
    toc: true
    toc_float: true
  word_document: default
  pdf_document:
    df_print: paged
    latex_engine: xelatex
always_allow_html: true
---

```{r setup, include=FALSE}
if (TRUE) {
  list.of.packages <- c("tidyverse", "haven", "here",
                      "kableExtra","rstudioapi","DescTools","qwraps2","data.table",
                      "stargazer","readxl","dplyr","rootSolve","ICC", "randomizr",
                      "multiwayvcov", "lmtest", "knitr","devtools", "Hmisc",
                      "compareGroups", "naniar")
  # sapply(list.of.packages, function(x) pacman::p_load(get(x)))

  new.packages <- list.of.packages[
    !( list.of.packages %in% installed.packages()[,"Package"] )
    ]
  if(length(new.packages)) install.packages(new.packages,
                                            repos = "https://cloud.r-project.org")

  lapply(list.of.packages, library, character.only = TRUE)
}

devtools::install_github('vikjam/pwrcalc')
## If you have trouble installing pwrcalc, you can download it manually by using the .tar.gz file provided. 
## To install it manually, go to "Packages", select "Install" and then "Package Archive" in the install menu. 

setwd(dirname(getActiveDocumentContext()$path)) 
knitr::opts_knit$set(root.dir = here())
knitr::opts_chunk$set(echo = T, eval = T, message = F, warning = F, fig.pos = "H", fig.height = 7, fig.width = 7)

print_code <- TRUE
```

# Endpoints
 
On RCTs especially in clinical trials, endpoints are prespecified in statistical analysis plans. They are primary and secondary endpoints in addition to safety and/or efficacy measures.

There are a few categories of endpoints.

- Continuous Endpoints: The outcome is described as continuous number. Usually, the mean and the standard deviation are used. If the distribution is skewed, the median can be displayed instead of the mean.

- Categorical Endpoints: The outcome can be described as binary (Yes or No). Usually, the percentage of each category is used.

- Composite Endpoints: Combine multiple categorical outcomes. Binary results are summarized as categorical variable.

- Count Endpoints: Non-normal distribution and discrete values. Histogram-like appearance.

- Survival (time-to-event) Endpoints: Assess the time to any recurrence of episode.


# Example 

Using the `balsakhi` dataset, provided with the package `pwrcalc`. We'll use this dataset to estimate the control mean:

```{r}
library(pwrcalc)
data(balsakhi)

# Let us view the `balsakhi `data that we have just loaded:
View(balsakhi)
```


## Continuous endpoint

If we want to know the percentage change from baseline in math score after the balsakhi program, we can explore the outcome as a continuous endpoint.
The variable `post_math` shows the absolute math score on post-test (integer).

```{r}
# Simple table of score comparison by group
balsakhi %>% 
  group_by(bal) %>% 
  summarise(mean_post_math=mean(post_math, na.rm=TRUE), sd_post_math=sd(post_math,na.rm=TRUE))


# Score change in math
bal_math <- balsakhi %>% 
  filter(!is.na(pre_math)&!is.na(post_math))
bal_math$diff_math <- bal_math$post_math - bal_math$pre_math
bal_math$pct_change_math <- bal_math$diff_math/bal_math$pre_math * 100

# visualize histogram

## new labels
labels <- c("0"="Control","1"="Treatment")

## plot
ggplot(data = bal_math, aes(x=pct_change_math))+
  geom_histogram(fill="white", color="black", binwidth = 50)+
  facet_grid(~bal, labeller = labeller(bal = labels))+ # change labels
  xlab("Percentage Change in Math from Pre-test")


```

## Categorical endpoint

If we want to know whether the percentage from baseline in passing rate of standard 1 math test decreased or i
The variable `postmath1std` shows if a student passed standard 1 math competencies in post-test (Passed = 1, Failed = 0).

```{r}
bal_math %>% 
  filter(!is.na(postmath1std)) %>% 
  count(bal,postmath1std) %>% 
  mutate(pct=100*n/sum(n))

# Tabulate with count
table(bal_math$postmath1std,bal_math$bal, dnn = c("Math1","Treatment"))

# Tabulate with proportion
prop.table(table(bal_math$postmath1std,bal_math$bal, dnn = c("Math1","Treatment")))

## If we want to shorten the digit
options(digits = 2)
prop.table(table(bal_math$postmath1std,bal_math$bal, dnn = c("Math1","Treatment")))
```

## Composite endpoint

If we are interested in whether respondents passed either standard 1 math test, standard 2 math test or standard 3 math test. Here we will generate a binary composite endpoint for this from the three variables in the `balsakhi` dataset.

```{r}
bal_math <- bal_math %>% 
  mutate(
    # Dichotomize the variable for math tests
    # Generate a combined binary endpoint for having any math tests pass
    combined = factor(
      ifelse(
        # Check for passing standard 1 math test
        postmath1std == 1 & 
          # ... and passing standard 2 math test
          postmath2std == 1 & 
          # ... and passing standard 3 math test
          postmath3std == 1,
        "All pass", 
        "At least one fail"
      ),
      levels = c("All pass", "At least one fail")
    )
  )

# Tabulate the new composite endpoint
table(bal_math$combined, useNA = "ifany")
```

# Hypothesis testing

Our interest is whether balsakhi program positively impacted the students academic performance.

- Null Hypothesis: $H_0: \mu_1 = \mu_2$

- Alternative Hypothesis (two-sided): $H_A: \mu_1 \neq \mu_2$

- Alternative Hypothesis (one-sided): $H_A: \mu_1 < \mu_2$

To test the hypothesis, we will estimate treatment effect (e.g. mean difference). Usually, confidence interval is 95\%.


Continuous

For continuous measure, if we find the changes from baseline within the groups are normally distributed, we can test whether the means are equal using a two-sample t-test.

`var.equal` argument is to specify whether we assume that the two variances, or standard deviations, are equal.

If we find the changes from baseline is not normally distributed (i.e. affected by skewness), we can use Wilcoxon Rank Sum test. Here the null hypothesis is the distribution of the two populations are the same. The alternative hypothesis is that there is a left or right translation shift.


Binary 

We can use the chi-squared test to test the null hypothesis of equal group proportions.


Important assumptions

We assume that the observations in the treatment groups are independent and that the randomization achieved similar patient characteristics between groups.



```{r}
# to show ratio in table,
createTable(baselines, show.ratio=TRUE)

# Odds Ratio (OR) is a measure of association between exposure and an outcome
```



